officedown template

### 2024-08-13 docs/verticality_future_mkd.Rmd
spatial varying relationships discussion

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall forests generally have greater foliage height diversity. These vegetation layers may act as a multi-floored apartment building that enables species to finely partition habitat [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness (*cite positive relationships between canopy height and species richness at local scales*). *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019]. However, canopy height and other metrics of vegetation structure do not consisently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

@@

We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality (Oliveira and Scheffers 2019). All environmental data was resampled to the resolution and projection of the assemblage data.

@@

The spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (\@ref(fig:mod-coefs), *supp table for model results*). Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality, though the impact of dry season precipitation on SES verticality was stronger (\@ref(fig:mod-coefs)). Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality. Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile mean and SES verticality, as well as SES verticality of mammals. This relationship was generally characterized by an initial increase in verticality with increasing maximum temperatures followed by a decline in verticality as temperatures increased further (*conditional effects plot*). Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship. Climate velocity negatively impacted bird and amphibian verticality and positively impacted mammal and reptile verticality. However, environmental factors were unable to explain extremely high and extremely low verticality as indicated by larger residuals at the ends of the verticality spectrum when spatial random effects were not included in estimates of assemblage verticality (*supp figure*).

@@

@@

>>>>

Vertical stacking of species is an emblematic signature of biologically rich regions and is associated with patterns of species richness from local to global scales [@macarthurPopulationEcologyWarblers1958; @macarthurBirdSpeciesDiversity1961; @oliveiraVerticalStratificationInfluences2019; @bashamVerticalStratificationPatterns2023a]. At local scales, species assemblages may be stratified across vertical resource and climate gradients from beneath the ground to the top of the canopy [@bashamVerticalStratificationPatterns2023a; @xingEcologicalPatternsProcesses2023]. Tall forests with high foliage hight diversity increase both the volume of habitat and the niche opportunities within the habitat, acting effectively as multi-floored apartments that enable species to vertically partition resources [@macarthurLimitingSimilarityConvergence1967; @macarthurBirdSpeciesDiversity1961; @larueDiversityVolumeRelationships2023]. By reducing competition, vertical stratification may thus act as a mechanistic link to positive relationships between canopy height and species richness [@wagnerRemotelySensedTree2023] (*cite positive relationships between canopy height and species richness at local scales*). *concept fig* This hypothesis has been supported by positive relationships between verticality, defined here as the average vertical position of species in an assemblage, and species richness [@oliveiraVerticalStratificationInfluences2019]. However, canopy height and other metrics of vegetation structure do not consisently impact species richness at regional to global scales, with notable declines in the strength of the relationship at higher latitudes where the diversity of arboreal species generally declines [@larueDiversityVolumeRelationships2023; @rollLinkingVertebrateSpecies2015; @oliveiraVerticalStratificationInfluences2019]. Examining the impacts of climatic and vegetative drivers of vertical stratification may thus add nuance to our understanding of the impact of canopy height on species richness.

@@

We obtained present and future global climate data from CHELSA v2.1 for mean daily maximum air temperature of the warmest month, mean daily minimum air temperature of the coldest month, temperature seasonality, precipitation of the wettest month, precipitation of the driest month, and precipitation seasonality [@kargerDataClimatologiesHigh2021; @kargerClimatologiesHighResolution2017]. Present climate data represents the time period 1981-2010 and future climate data represents an ensemble of climate scenarios for SSP585 for the time period 2071-2100. Vegetation data included canopy height [@langHighresolutionCanopyHeight2023], vegetation density [@crowtherMappingTreeDensity2015], and vegetation complexity (the product of canopy height and vegetation density). Additionally, we included historical climate velocity, which represents the average rate at which thermal isotherms have shifted between the last glacial maximum and present climate conditions [@sandelInfluenceLateQuaternary2011] and has been shown to influence global patterns of verticality [@oliveiraVerticalStratificationInfluences2019]. All environmental data was resampled to the resolution and projection of the assemblage data.

@@

The spatial GLMMs indicated that verticality was limited by low precipitation across taxa, while impacts of high maximum temperatures, low minimum temperatures, and canopy height varied (\@ref(fig:mod-coefs), *supp table for model results*). Across taxa, precipitation of the driest and wettest months positively influenced mean and SES verticality (\@ref(fig:mod-coefs)). Minimum temperature of the coldest month positively impacted mammal and reptile verticality and had negligible effects on bird and amphibian verticality. Maximum temperatures of the warmest month exhibited a quadratic relationship with bird and reptile mean and SES verticality, as well as SES verticality of mammals. Mean verticality of mammals exhibited a positive linear relationship with maximum temperature and amphibian verticality exhibited no relationship. These patterns resulted in declines in verticality at both extremely high and low maximum temperatures in regions without high dry season precipitation for reptiles and amphibians, while verticality increased with temperature across a range of dry season precipitation for birds and mammals (*see conditional effect plots*). Climate velocity negatively impacted bird and amphibian verticality and positively impacted mammal and reptile verticality. However, environmental factors were unable to explain extremely high and extremely low verticality as indicated by larger residuals at the ends of the verticality spectrum when spatial random effects were not included in estimates of assemblage verticality (*supp figure*).

@@

Our results indicate globally consistent macroscale patterns of verticality across taxa that are characterized by declining verticality toward higher latitudes and correspond with previous studeis that are more limited in taxonomic or spatial extent [@bashamVerticalStratificationPatterns2023a; @oliveiraVerticalStratificationInfluences2019]. Though climate conditions influenced verticality, they were neither able to predict extremely high or low verticality nor spatial variation in the influence of canopy height on verticality. Furthermore, `r word_comment(highlight = "Furthermore", comment = "Try bootstrapping method to compare slopes")` regions exhibitin positive relationships between canopy height and verticality did not consistently exhibit stronger positive relationships between verticality and species richness. Our results thus do not support our hypothesis that climtatic conditions that support high assemblage verticality lead to positive impacts of canopy hiehg on verticality and positive associations between verticality and species richness. Vertical stratification across increasingly tall and multi-layered forests may therefore not be a primary mehcanism driving positive relationships between canopy height, verticality, and species richness at global scales. Rather, we propose that species packing and evolutionary patterns may explain our unexpected results. The ways in which climate change will impact verticality and its relationship with species richness is thus difficult to predict.

Climate influences global patterns of verticality because more extreme climates in teh canopy, including increased exposure to hot, cold, and dry conditions, pose physiological challenges to living in arboreal habitats [@oliveiraVerticalStratificationInfluences2019]. Water availabitliy therefore facilitates verticality, while temperatures at both hot and cold extremes and awter scaracity generally limit verticality. For exmaple, more severe water limitation in the canopy than the understory [@scheffersIncreasingArborealityAltitude2013] may contribute to observed declines in arboreal life straties outside moist forests in tropical regions. However, our models indicated nuances in the relative influences of different drivers across taxa.

Impacts of precipitation were particularly pronounced for amphibian verticality. In tropical wet forests, evapotranspiration allows amphibian body temperatures to remain below critical thermal maxima when exposed to high temperatures because water is consitently available (*cite*). However, in regions that experience pronounced dry seasons, amphibians behaviourally thermoregulate to avoid desiccation by increasing their use of understory habitats, which reduces assemblage verticality [@bashamVerticalStratificationCollapses2020]. Lower precipitation poses a similar constraint to arboreality for amphibians in dry regions at higher latitudes where basking to increase body temperatures can quickly lead to desccation witout adequate moisture availability or physiological defenses [@sundayThermalsafetyMarginsNecessity2014; @tracyNotJustSmall2010]. *Birds were also influenced by precipitation, likely because they are dependent on the availability of standing water in teh canopy and water content of leaves.* `r word_comment(highlith = "leaves", comment = "need to look more into impacts of drought on birds - see https://onlinelibrary.wiley.com/doi/full/10.1002/ece3.10781")`

For mammals and reptiles, our modesl indicated that minimum cold temperatures had the greatest impact on verticality. In boreal forests, cold and windy conditions in the canopy pose large energetic costs to endotherms and can cause body temperatures to drop below critical thermal minima in ectotherms [@bymanEnergeticCostsWinter1988; @sundayThermalsafetyMarginsNecessity2014]. Species are therefore largely constratined to terrestrial or fossorial habiats, which are buffered from minimum temperature extremes [@defrenneGlobalBufferingTemperatures2019]. Yet despite the limiting or facilitative impacts that climate can have on verticality, climate conditions in combination with vegetation structure failed to predict extremely high and low verticality when the spatial random field was excluded from predictions, suggesting other factors contribute to unexplained variation.

## Spatially varying relationships between canopy height, verticality, and richness

Our models indicated that canopy height has a variable and often insignificant impact on verticality, even across much of the tropics where tall, multi-layered forests would be expected to promote vertical expansion and high diversity of arboreal species [@burnsGriddedGEDIVegetation2024; @fengForestCanopyHeight2020; @rollLinkingVertebrateSpecies2015]. These results align with Basham et al. [-@bashamVerticalStratificationPatterns2023a], who similarly found an insignificant impact of canopy height on verticality within the tropics. Interestingly, previously reported spatial variation in the impact of canopy height on taxonomic and functional richness mirror, to some extent, spatial variation in the impacts of canopy height on verticality [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020; @coopsDisentanglingVegetationClimate2018]. For example, canopy height positively impacts verticality and species richness of amphibians in Madagascar, the Andes, Central America, Southern India, and New Guinea [@rollLinkingVertebrateSpecies2015]. Spatially concurrent positive impacts for birds occur in the Himalayas, New Guinea, and southeastern Australia, while for mammals they occur in northwestern Mexico and eastern Madagascar [@coopsDisentanglingVegetationClimate2018; @rollLinkingVertebrateSpecies2015]. Additionally, concurrent negative impacts for mammals occur in parts of the Amazon basin (\@ref(fig:svc-sig)) [@rollLinkingVertebrateSpecies2015; @fengForestCanopyHeight2020]. Given spatial similarities in relationships between canopy height and both verticality and richness, we expected stronger positive correlations between verticality and richness in regions where canopy height positively impacted verticality. Yet, our results did not consistently support this hypothesis.

A niche expansion model of community assembly characterizes the hypothesis that taller canopies with greater foliage height diversity allow greater vertical stratification and subsequent increases in verticality and species richness [@macarthurBirdSpeciesDiversity1961; @macarthurPatternsSpeciesDiversity1965]. The niche expansion model posits that species richness increases when additional species occupy novel habitat or resource space [@pellissierNichePackingExpansion2018; @pigotFunctionalTraitsReveal2016]. Thus, arboreal species may be added to the assemblage by occupying the novel habitat space that becomes available as canopy height increases, producing a stronger positive relationship between verticality and richness in regions where canopy height has a positive influence on verticality. This pattern was evident for birds and amphibians. However, this three-way positive relationship did not occur in the most diverse regions, including much of the tropics. Instead, it was largely restricted to areas where relatively rapid increases in vegetation structure from high to low elevations or dry to wet forests (e.g., Madagascar) may facilitate the addition of species outside the habitat space occupied by species in assemblages of less productive habitats [@verschuylEffectForestStructure2008; @pellissierNichePackingExpansion2018]. Yet in regions where canopy height did not have significant positive impacts on verticality, positive correlations between verticality and richness remained for birds and amphibians and increased in strength for mammals and reptiles, suggesting vertical stratification (i.e., niche expansion) may not be the dominant mechanism causing positive relationships between canopy height, verticality, and species richness at large spatial scales.

Rather than expansion across tall forests, positive associations between verticality and species richness could be driven by niche packing. Niche packing occurs when smaller niche spaces or greater niche overlap allow more species to co-exist within a habitat [@pigotFunctionalTraitsReveal2016; @pellissierNichePackingExpansion2018]. Specialization on specific resources within the canopy over evolutionary time could thus increase arboreal species diversity without increasing canopy height [@fengForestCanopyHeight2020; @pellissierNichePackingExpansion2018]. Furthermore, evolutionary rates of arboreal lineages are generally faster than those of terrestrial and fossorial lineages [@cyriacDiggingTheirOwn2018; @liWhatDrivesDiversification2022; @moenMicrohabitatClimaticNiche2017]. In the tropics where climate conditions support arboreality, the combination of niche packing and rapid evolutionary rates of arboreal species could lead to a disproportionately high number of arboreal species, increasing verticality relative to what would be expected by climate conditions alone and causing positive relationships between verticality and richness in regions where canopy height has insignificant or even negative effects on verticality.

@@

```{r fig.id="svc-sig", fig.lp="ed", fig.cap="Areas in which the spatially varying coefficient for canopy height had a significant positive or negative effect on SES or mean verticality of birds, mammals, reptiles, and amphibians.", fig.height=143.9, fig.width=160}

include_graphics("./../figures/supp_figs/svc_canopyHeight_sig_link.png", error = F)
```
